% ==============================================================================
% MAIN DOCUMENT
% Forecasting International Migration to North Dakota: A Multi-Method Analysis
% ==============================================================================
% Master file that compiles all sections of the journal article
% ==============================================================================

\input{preamble}
\IfFileExists{build_info.tex}{\input{build_info}}{
    \newcommand{\BuildVersion}{DEV}
    \newcommand{\BuildTimestampUTC}{}
}
\IfFileExists{derived_stats.tex}{\input{derived_stats}}{
    \newcommand{\PEPLatestYear}{2024}
    \newcommand{\PEPShareYearStart}{2010}
    \newcommand{\PEPShareYearEnd}{2024}
    \newcommand{\PEPShareNYears}{15}
    \newcommand{\NDShareUSIntlMigPctLatest}{0.18}
    \newcommand{\NDShareUSPopPctLatest}{0.23}
    \newcommand{\NDShareUSIntlMigPctMean}{0.17}
    \newcommand{\NDShareUSPopPctMean}{0.23}
    \newcommand{\PEPContextWindowYears}{5}
    \newcommand{\PEPPriorFiveYearStart}{2015}
    \newcommand{\PEPPriorFiveYearEnd}{2019}
    \newcommand{\PEPLastFiveYearStart}{2020}
    \newcommand{\PEPLastFiveYearEnd}{2024}
    \newcommand{\PEPLastTenYearStart}{2015}
    \newcommand{\PEPLastTenYearEnd}{2024}
    \newcommand{\NDNaturalIncreaseMeanPriorFive}{4,724}
    \newcommand{\NDNaturalIncreaseMeanLastFive}{2,217}
    \newcommand{\NDDomesticMigMeanLastTen}{-1,392}
    \newcommand{\NDDomesticMigNegYearsLastTen}{8}
    \newcommand{\NDDomesticMigYearsLastTen}{10}
}

\begin{document}

% ------------------------------------------------------------------------------
% Internal Draft Cover Sheet
% ------------------------------------------------------------------------------
\thispagestyle{empty}
\begin{center}
{\Large\bfseries WORKING DRAFT}\\[0.5em]
{\normalsize Internal review copy}\\[1.5em]
\begin{minipage}{0.85\textwidth}
\small
\noindent Please do not redistribute without contacting the author. Subject to public disclosure upon request.\\[0.75em]
\noindent Build: v\BuildVersion\ (\BuildTimestampUTC).
\end{minipage}
\end{center}
\newpage

% ------------------------------------------------------------------------------
% Title Page
% ------------------------------------------------------------------------------

\title{Forecasting International Migration to North Dakota: \\
A Multi-Method Empirical Analysis}

\author{
    Nigel Haarstad \\[0.25em]
    \small State Demographer / Senior Economic Data Analyst \\
    \small North Dakota Department of Commerce --- State Data Center \\
    \small nhaarstad@nd.gov
}

\date{\today}

\maketitle
\begingroup
\renewcommand{\thefootnote}{}
\footnotetext{Corresponding author: nhaarstad@nd.gov. Author background: The author is State Demographer/Senior Economic Data Analyst at the North Dakota Department of Commerce and holds a PhD in Communication (University of Kentucky, 2016). The views expressed are those of the author and do not necessarily represent the views of the North Dakota Department of Commerce. Funding: No external funding was received. Conflicts of interest: The author declares no conflicts of interest. Data/code availability: Available upon request.}
\addtocounter{footnote}{-1}
\endgroup

% ------------------------------------------------------------------------------
% Abstract
% ------------------------------------------------------------------------------

\begin{abstract}
International migration is numerically small in many U.S.\ states yet demographically decisive in small-population, non-gateway states, where low fertility, population aging, and domestic out-migration leave migration as the primary swing factor in population change. Forecasting these flows is unusually difficult because arrivals are often ``lumpy''---shaped by discrete refugee resettlement cohorts and employer-specific recruitment---and because key levers are federal policy and local resettlement capacity rather than stable long-run trends. This study develops and applies a multi-method empirical framework for forecasting net international migration to North Dakota, a state of roughly 797,000 residents that accounts for about \NDShareUSIntlMigPctMean\% of U.S.\ net international migration on average over \PEPShareYearStart{}--\PEPShareYearEnd{}---below its \NDShareUSPopPctMean\% share of the U.S.\ population.

Using the Census Bureau Population Estimates Program (\PEP) net international migration component as the forecasting target (2010--2024) and triangulating with Refugee Processing Center arrivals (FY2002--FY2024), DHS lawful permanent resident admissions, and \ACS{} foreign-born stocks, we combine descriptive concentration measures, time-series diagnostics, cross-state benchmarking, gravity-style allocation models, machine-learning diagnostics, and difference-in-differences analysis of policy-associated divergence, alongside wave-duration modeling and Monte Carlo scenario simulation.

Results show that North Dakota's migration profile is refugee-dominated and origin-specific: location quotients exceed 40 for Liberia and exceed 20 for several other African origins. Annual net international migration is highly volatile (coefficient of variation 82.5\%) and the series is approximately integrated, implying that departures from historical averages can persist rather than rapidly mean-revert. Structural break analysis identifies major shifts in migration dynamics in 2020 and 2021 associated with the COVID-19 pandemic ($F = 16.01$, $p < 0.001$). Aggregate break tests detect no statistically significant 2017 shift, yet nationality-level difference-in-differences estimates a policy-associated treated--control divergence, with estimated refugee arrivals from Travel Ban-affected origins roughly 75\% lower than those from unaffected origins in the first full post years (conventional $p = 0.031$). Given mixed pre-trends (parallel-trends diagnostics) and only seven treated nationalities, we interpret this pattern as descriptive policy-associated divergence rather than a precise causal effect magnitude, highlighting how compositional divergence can be muted in state totals. Cross-sectional allocation estimates suggest diaspora associations attenuate sharply once population mass is controlled.

Scenario projections through 2045 yield a median net international migration of 9,079 with a conservative 95\% uncertainty envelope of 3,360--14,774. These results argue for scenario-based planning and transparent uncertainty quantification when forecasting migration in small, policy-sensitive states.
\end{abstract}

\vspace{1em}

\noindent\textbf{Keywords:} international migration, forecasting, North Dakota, causal inference, demographic projection, refugee resettlement, Great Plains

\vspace{0.5em}

\noindent\textbf{JEL Codes:} J11, J61, C22, C53, R23

\newpage

% ------------------------------------------------------------------------------
% Table of Contents (optional, comment out for submission)
% ------------------------------------------------------------------------------
% \tableofcontents
% \newpage

% ------------------------------------------------------------------------------
% Main Sections
% ------------------------------------------------------------------------------

\input{sections/01_introduction}

\input{sections/02_data_methods}

\input{sections/03_results}

\input{figures/figure_captions}

\input{sections/04_discussion}

\input{sections/05_conclusion}

% ------------------------------------------------------------------------------
% References
% ------------------------------------------------------------------------------

\newpage
\bibliographystyle{apalike}
\bibliography{references}

% ------------------------------------------------------------------------------
% Appendix
% ------------------------------------------------------------------------------

\newpage
\appendix
\input{sections/06_appendix}

% ==============================================================================
% END DOCUMENT
% ==============================================================================

\end{document}
