% ==============================================================================
% CONCLUSION
% Forecasting International Migration to North Dakota: A Multi-Method Analysis
% ==============================================================================

\section{Conclusion}
\label{sec:conclusion}

% ==============================================================================
% 5.1 SUMMARY OF CONTRIBUTIONS
% ==============================================================================

\subsection{Summary of Contributions}
\label{subsec:contributions}

This study offers three principal contributions to the migration forecasting literature. First, it provides the first comprehensive multi-method empirical analysis of international migration to a small U.S. state. The nine-module analytical framework---spanning descriptive statistics, time series analysis, panel regression, gravity models, machine learning, causal inference, and duration analysis---demonstrates that rigorous demographic analysis remains feasible even when small populations and short time series strain conventional asymptotic methods. The multi-method design explicitly trades depth within any single paradigm for robustness across paradigms, generating convergent findings that warrant greater confidence than single-method analyses could support.

Second, the analysis demonstrates novel applications of causal inference methods to state-level migration policy evaluation. The difference-in-differences estimate indicates a policy-associated divergence of approximately 75\% in refugee arrivals from Travel Ban-affected countries (conventional $p = 0.031$), but mixed parallel-trends diagnostics and small treated-cluster count imply that the causal magnitude is uncertain; we therefore treat the estimate as descriptive policy-sensitivity evidence. A synthetic comparator provides descriptive benchmarking for North Dakota without causal interpretation, and a shift-share index establishes first-stage relevance linking refugee-driven shocks to state migration. Together, these elements extend the migration policy evaluation toolkit beyond the national-level and gateway-state analyses that dominate existing literature while keeping causal claims appropriately bounded.

Third, the study develops a rigorous framework for uncertainty quantification in long-range migration forecasts. The scenario analysis distinguishes forecast uncertainty (parametric imprecision within an immigration policy regime) from scenario uncertainty (unknowable future policy choices), while Monte Carlo simulation propagates parameter uncertainty through projection models to generate probability distributions rather than misleadingly precise point estimates. The resulting 95\% uncertainty envelope spanning a factor of 4.4 for 2045 migration honestly characterizes the structural uncertainty inherent in this forecasting domain.

% ==============================================================================
% 5.2 KEY TAKEAWAYS
% ==============================================================================

\subsection{Key Takeaways}
\label{subsec:takeaways}

Three empirical conclusions emerge with particular force from the analysis. First, North Dakota's international migration is refugee-dominated and policy-sensitive. The Location Quotient analysis reveals extreme overrepresentation of African and Middle Eastern origins---products of federal resettlement decisions rather than economic migration chains---and the Travel Ban analysis provides evidence of large policy-associated divergence in refugee arrivals from affected origins. Demographic planners must recognize this policy dependence as a fundamental feature of the forecasting problem.

Second, the high volatility of small-state migration requires scenario-based planning rather than point forecasting. The coefficient of variation of 82.5\%, the random-walk baseline characterization, and the dramatic 2020 structural break collectively indicate that migration levels can shift by amounts approaching the mean itself. The deterministic scenarios span 2,138 to 7,048 migrants by 2045, while the wave-adjusted 95\% uncertainty envelope spans 3,360 to 14,774. The true trajectory likely evolves within this wide range and remains heavily conditioned by federal policy decisions.

Third, diaspora associations in North Dakota appear modest but present. The diaspora elasticity of 0.14---below estimates from gateway state and cross-national studies---suggests that refugee resettlement has not yet generated the self-reinforcing migration chains observed in more established immigrant communities. This finding implies both vulnerability (migration depends heavily on external policy rather than internal momentum) and opportunity (investments in community development may strengthen diaspora associations over time).

% ==============================================================================
% 5.3 FUTURE RESEARCH DIRECTIONS
% ==============================================================================

\subsection{Future Research Directions}
\label{subsec:future_research}

Several avenues warrant further investigation. Extension of this analytical framework to other Great Plains states would assess whether North Dakota's patterns generalize across the region or reflect state-specific idiosyncrasies. States including South Dakota, Nebraska, and Minnesota share demographic challenges and refugee resettlement participation, potentially enabling comparative analysis of network development and policy sensitivity across contexts.

Incorporation of county-level dynamics would illuminate the geographic distribution of migration within North Dakota. Urban centers including Fargo and Grand Forks receive the majority of refugee arrivals, but secondary migration to smaller communities may follow different patterns. County-level analysis could inform local planning and identify communities where service capacity constraints bind.

Integration with labor market outcomes would address whether international migration alleviates the workforce shortages that motivate policy interest in immigration to rural regions. Linking migration data to employment outcomes, wage dynamics, and industry-level labor supply could quantify the economic contributions of immigrant populations and inform workforce development investments.

Finally, real-time forecasting with updated data would enable the scenario framework developed here to evolve as new information becomes available. The post-COVID recovery trajectory remains incompletely characterized given data limitations; ongoing updates could refine the 2020 structural break assessment and improve medium-term projections as post-COVID migration dynamics stabilize.

\vspace{1em}

International migration to North Dakota occupies a distinctive position in American demography: it is small in absolute terms, volatile over time, concentrated in humanitarian categories, and sensitive to federal policy decisions. Understanding these patterns requires analytical frameworks suited to their inherent complexity and uncertainty. This study has sought to provide such a framework by combining methodological rigor with honest acknowledgment of forecasting limits. The contribution lies not in resolving uncertainty but in characterizing it accurately, enabling demographic planners to develop robust strategies that perform reasonably across plausible futures rather than optimizing for misleadingly precise point forecasts.
